Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-07 paper/ms.tex
update ms

To date, there are four known stellar ensembles in the Kepler field that have
yielded a total of fourteen transiting planets.
Our analysis blindly recovers each of these planets, as well as their ages.

@@

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age
increases, because the two techniques are independent.
The region of greatest overlap between rotation and lithium ages in
the spaces of $P_{\rm rot}$, EW$_{\rm Li}$, and $T_{\rm eff}$ is
dictated by the timescale for lithium depletion: for stars with
temperatures between 4500-5500\,K with ages $\lesssim$500\,Myr, we can
hope to obtain independent two-sided constraints
\citep{Bouma_2023,Jeffries_2023}.

@@

The kinematic distinction between thin and thick disk stars helps connect
the Kepler stars to the broader literature focused
on the history of star formation in the Galaxy.  On extragalactic
scales, the star formation rate in spiral galaxies with mass
comparable to the Milky Way peaked $\approx$10\,Gyr ago, and has since
decreased by about an order of magnitude
\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  For our own
galaxy, many investigators have worked to measure similar star
formation histories, through both CMD fitting of resolved stellar
populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
and also by modelling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  
\citet{2019A&A...624L...1M} and \citet{2019ApJ...878L..11I} for
instance focused on stars (and white dwarfs) within a few hundred
parsecs, and both reported a local maximum in the galactic star
formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R} focused on
stars in a wider 2\,kpc bubble, and reported {\it three} local maxima
in the thin disk's star formation history 5.7, 1.9, and 1.0\,Gyr ago,
which they associate with pericenter passages of the Sagittarius satellite
galaxy.

@@

Other age indicators may help in verifying the ages of
the $\lesssim$3\,Gyr stars that are the focus of this work.
Specifically, chromospheric emission in the
X-ray, \ion{Ca}{2} HK line, and the UV can serve as an age tracer
\citep{2008ApJ...687.1264M,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.  The dynamo
converts kinetic energy into magnetic energy, which is emitted through these chromospheric
pathways.
We did not attempt to incorporate these indicators into this study
due to concerns regarding sensitivity, homogeneity, and
the question of whether they in fact provide age
information that is truly independent from rotation.

A broader question, beyond the scope of this work, concerns the ages of
the older $\gtrsim$3\,Gyr stars that we do not treat.
For age-dating models based only on rotation rates, or else
on a combination of rotation and spin-modulation amplitude
\citep{2023ApJ...952..131M}, an important challenge is that the detectability
of the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
is at the limits of the current data \citep{2022ApJ...937...94M}.
In the ``old star'' regime, methods that use both the evolution of
stellar luminosity and stellar rotation \citep{Angus_2019} seem the most
capable of providing useful age constraints for single stars.
Incorporating kinematics \citep{2021AJ....161..189L} seems likely to remain
most useful for stellar ensembles, due to the stochastically of kinematic heating.

\subsection{Future Directions}
{\it Occurrence Rates}: Some interesting
age-evolution trends have been predicted, and can probably be teased
from our data.  
Kelvin-Helmholtz cooling for mini-Neptunes should scale $R_{\rm p}$ as
$t^{-0.1}$ \citep{Gupta_2019}.
The photoevaporation desert has been predicted to be ``carved'' over the timescale
for atmospheric loss \citep{Owen_Lai_2018}.
Mini-Neptunes should be converted into super-Earths \citep{Rogers_2021}.
And the boundary of the super-Earth population may correspondingly move up in size over
time \citep{David_2021}.
I'd prefer to split the two papers.

{\it Field star ages from other surveys}:
In this study we have focused on deriving rotation and lithium-based ages for the Kepler
planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
Rotation-based age dating of their discoveries might be useful for the effort
of identifying young exoplanets, and also for the broader issue of understanding
their evolution.

{\it Searches for new associations}:
Some of the youngest systems in Table~\ref{tab:planets} have been
connected to their birth clusters.
Some have not.
A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
yet to be performed.
The ideal approach for such a search would be to combine spatial and kinematic information
from Gaia with rotation measurements from TESS.

%   {\it }
%   
%   
%   {\it Early carving of the photoevaporation desert}
%   \citep{Owen_Lai_2018}
%   
%   {\it Rapidly decreasing abundance of big puffy planets}
%   Kepler not enough?
%   
%   {\it Mini-Neptunes turning into Super-Earths}
%   \citep[e.g.][]{Rogers_2021}
%   
%   {\it Boundary of the Super-Earth population moving up in time}
%   \citep{David_2021}

@@

%\clearpage

>>>>

There are currently four known stellar ensembles in the Kepler field that have
yielded a total of fourteen transiting planets.
Our analysis blindly recovers each of these planets, as well as their ages.

@@

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age
increases, because the two techniques are independent.
The region of greatest overlap between rotation and lithium ages in
the spaces of $P_{\rm rot}$, EW$_{\rm Li}$, and $T_{\rm eff}$ is
dictated by the timescale for lithium depletion: for stars with
temperatures between 4500-5500\,K with ages $\lesssim$1\,Gyr, we can
hope to obtain independent two-sided constraints
\citep{Bouma_2023,Jeffries_2023}.

@@

The kinematic distinction between thin and thick disk stars helps connect
the Kepler stars to the broader literature focused
on the history of star formation in the Galaxy.  On extragalactic
scales, the star formation rate in spiral galaxies with mass
comparable to the Milky Way peaked $\approx$10\,Gyr ago, and has since
decreased by about an order of magnitude
\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  For our own
galaxy, many investigators have worked to measure similar star
formation histories, through both CMD fitting of resolved stellar
populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
and also by modelling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  
\citet{2019A&A...624L...1M} and \citet{2019ApJ...878L..11I} for
instance focused on stars (and white dwarfs) within a few hundred
parsecs, and both reported a local maximum in the galactic star
formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R} focused on
stars in a wider 2\,kpc bubble, and reported three local maxima
in the thin disk's star formation history 5.7, 1.9, and 1.0\,Gyr ago,
which they associate with pericenter passages of the Sagittarius satellite
galaxy.

@@

Other age indicators may help in verifying the ages of
the $\lesssim$3\,Gyr stars that are the focus of this work.
Specifically, chromospheric emission in the
X-ray, \ion{Ca}{2} HK line, and the UV can serve as an age tracer
\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.  The dynamo
converts kinetic energy into magnetic energy, which is emitted through these chromospheric
pathways.
We did not attempt to incorporate these indicators into this study
due to concerns regarding sensitivity, homogeneity, and
the question of whether they in fact provide age
information that is truly independent from rotation.

A broader question, beyond the scope of this work, concerns the ages of
the older $\gtrsim$3\,Gyr stars that we do not treat.
For age-dating models based only on rotation rates, or else
on a combination of rotation and spin-modulation amplitude
\citep{2023ApJ...952..131M}, an important challenge is that the detectability
of the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
is at the limits of the current data \citep{2022ApJ...937...94M}.
In the ``old star'' regime, methods 
using either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else
using both the evolution of
stellar luminosity and stellar rotation \citep{Angus_2019} seem the most
capable of providing useful age constraints for single stars.
Incorporating kinematics \citep{2021AJ....161..189L} seems likely to remain
most useful for stellar ensembles, due to the stochastically of kinematic heating.

\subsection{Future Directions}
{\it Occurrence Rates}: Some interesting
age-evolution trends have been predicted, and can probably be teased
from our data.  
Kelvin-Helmholtz cooling for mini-Neptunes should scale $R_{\rm p}$ as
$t^{-0.1}$ \citep{Gupta_2019}.
The photoevaporation desert has been predicted to be ``carved'' over the timescale
for atmospheric loss \citep{Owen_Lai_2018}.
Mini-Neptunes should be converted into super-Earths \citep{Rogers_2021}.
And the boundary of the super-Earth population may correspondingly move up in size over
time \citep{David_2021}.
We're going to split the two papers.

{\it Field star ages from other surveys}:
In this study we focused on deriving rotation and lithium-based ages for the Kepler
planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
Rotation-based age dating of their discoveries could be useful for the effort
of identifying young exoplanets, and also for the broader issue of understanding
their evolution.

{\it Searches for new associations}:
Some of the youngest systems in Table~\ref{tab:planets} have been
connected to their birth clusters.
Others have not.
A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
yet to be performed.
The ideal approach for such a search would be to combine spatial and kinematic information
from Gaia with rotation measurements from TESS.
Given that the decoherence time of young clusters is $\sim$100\,Myr, one would expect that
the youngest ``field stars'' may have young neighbors that are relatively easy to
identify.

@@
